\section{Acknowledgements}
We would like to thank Prof.\ Dr.\ R\"udiger Valk and Dr.\ Daniel Moldt
from the University of Hamburg for interesting discussions, help, 
and encouraging comments. 

We would also like to thank 
S\"onke R\"olke,
Dennis Schmitz and
Martin Wincierz
for their work during the preparation of this
release.
We would like to thank J\"orn Schumacher
for the prototype of the plug-in system (2.0), Benjamin Schleinzer
for his work during the preparation of former releases (2.1-2.2) and 
Berndt M\"uller who %%Farwer 
has been of great help with respect to previous
Renew releases for Mac OS ($\leq{}$2.0).
Some nice extensions of Renew were suggested or programmed
by Michael K\"oh\-ler and Heiko R\"olke.

We are indebted to the authors of various freeware libraries, namely
Mark Donszelmann, %%% freehep vectorgraphics
Erich Gamma, %%% JHotDraw
Doug Lea, %%% collections package (still used in FS)
David Megginson, %%% SAX
Bill McKeeman %%% Java Grammar
and Sriram Sankar. %%% JavaCC
%Marc Prud'hommeaux (JLine1), Jason Dillon, Guillaume Nodet, Colin Jones (JLine2)

Dr.\ Maryam Purvis, Dr.\ Da Deng, and Selena Lemalu 
from the Department of Information Science
(\texttt{http://infosci.otago.ac.nz/}), University of
Otago, Dunedin, New Zealand,
kindly aided us in the translation of parts of the documentation
and are involved in an interesting application project.

Valuable contributions and suggestions were made by
students and scientific workers at the University of Hamburg, most notably
Hannes Ahrens,
Tobias Betz, 
Jan Bolte,
Lars Braubach,
Timo Carl,
Dominic Dibbern,
Friedrich Delgado Friedrichs,
Matthias Ernst, 
Max Friedrich,
Daniel Friehe,
Olaf Gro\ss ler, 
Julia Hagemeister,
Sven Heitsch,
Marcin Hewelt,
Jan Hicken,
Thomas Jacob, 
Andreas Kanzlers,
Lutz Kirsten,
Michael K\"oh\-ler, 
Till Kothe,
Annette Laue, 
Matthias Liedtke, 
Marcel Martens, 
Klaus Mit\-rei\-ter,
Konstantin M\"ollers,
Eva M\"uller,
Jens Nor\-gall,
Sven Offermann,
Felix Ortmann,
Martin Pfeiffer, 
Alexander Pokahr,
Tobias Rathjen,
Christian R\"oder,
Heiko R\"olke, 
Benjamin Schleinzer,
Jan Schl\"uter,
Marc Sch\"on\-berg,
J\"orn Schumacher,
Michael Simon,
Fabian Sobanski, 
Volker Tell,
Benjamin Teuber,
Thomas Wagner,
Matthias Wester-Ebbinghaus, 
and Eberhard Wolff. 



We would like to thank the numerous users of Renew who provided
hints and constructive criticism. They helped greatly in improving the
quality of the code and the documentation. In particular, 
we would like to name Alun Champion and Zacharias Tsiatsoulis.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "renew.tex"
%%% End: 
